---
title: "DataVis Class: Final Project Report"
author: "Group 'Data Wizards': Aishe Memetova, Haotian Bu, Jon White, Vitalii Zakhozhyi"
output: html_document
---
<h1 align="center" style="color:blue">Exploring cash use as a transit fare payment in Seattle</h1>

Our group is interested in cash use as fare payment method in the Greater Seattle area.

Using the Seattle Sound Transit Data of October-November 2015, we will try to answer the following questions:
1) Are cash payments still have a big stake in transit fare types among riders in Seattle?
2) What type of fare is the most common given different purpose of trips?
3).....

**Loading neccessary packages, calling dataset and preparing it for analysis**

First we need to call the libraries to be able to use the packages 

```{r load library, results='hide', warning=FALSE, message=FALSE}
library(ggplot2)
library(scales)
library(dplyr)
```
Then we need to call the data we want to analyze. In our case we call the csv file

```{r import data, results='hide', warning=FALSE, message=FALSE}
linkdata <- read.csv("https://raw.githubusercontent.com/karllovepolicy/Data_Visual_Final_Tutorial/master/FinalData.csv", stringsAsFactors=FALSE)
```

**Finalizing the data frame for ggplot**

We will need to get rid of missing values so that they do not distort our analysis. Then we will need to separate data we are interested to analyze (subsetting the data), and calculate the frequences

```{r clean, warning=FALSE, message=FALSE}
linkdata_clean <- linkdata %>%
  filter(faretype != "#NULL!")

absol_ft=table(linkdata_clean$faretype, exclude = 'nothing')
names(absol_ft)[2]='Other'
ft_freq=prop.table(absol_ft)*100
ft_freq_df=as.data.frame(ft_freq) 
names(ft_freq_df)=c("faretype","pct")
ft_freq_df_ord=ft_freq_df[order(-ft_freq_df$pct), ]
```

**Building a bar plot using ggplot**

Now we can build a bar plot to visualize how frequently each of the payment methods is used. Since our primary focus is cash, we want to highlight it by using the red color for the bar which represents cash. 
```{r barplot}
cols <- ifelse(ft_freq_df_ord$faretype == "Cash", "red","gray50")

base = ggplot(ft_freq_df_ord, aes(faretype, pct))
ft_plot1 = base + geom_bar(stat="identity", fill=cols, width = 0.75)
ft_plot2 = ft_plot1 + labs(title='Do riders still use cash to pay a transit fare in Seattle?', #to label the graph 
                           x ='Type of fare',
                           y = 'Percentage of riders',
                           caption = 'Source: Sound Transit 2015 Onboard Survey Data')
ft_plot3 = ft_plot2 + coord_flip() +
                      scale_y_continuous(limits=c(0,55),
                                         breaks = seq(0, 55, by = 5), 
                                         labels=scales::unit_format(suffix = '%')) +
                      scale_x_discrete(limits=ft_freq_df_ord$faretype)
ft_plot4 = ft_plot3 + geom_text(aes(y = pct,
                                label = paste0(round(pct,2), '%')), 
                                hjust= -0.15, 
                                size = 3)
ft_plot5 = ft_plot4 + theme(panel.background = element_rect(fill = "white", color = "black"),
                            plot.title = element_text(size = 15, hjust = 0.5),
                            plot.caption = element_text(hjust = 1, vjust = -1.2),
                            axis.title.y = element_text(size = 12, hjust = 0.5, vjust = 1.75),
                            axis.title.x = element_text(size = 12, hjust = 0.5, vjust = -1.25))
ft_plot5
```


###################### BIVAR #####################   

Now, we will try to answer what type of fare is the most common given different purpose of trips.


**Cleaning the data and creating the proportion table**

We will first need to clean the data so that there are no missing values

```{r linkdataclean}
linkdata_clean <- linkdata %>%
  filter(faretype != "#NULL!") %>%
  filter(purpose != "#NULL!")

linkincomefare=table(linkdata_clean$faretype, linkdata_clean$purpose)%>%
         prop.table(margin = 2)%>%   # 2 is % by column
         "*"(100)%>%
         round(3)
```

**Creating the dataframe for building the plot**

Same as with the barplot, we will need to create a data frame

```{r dflink}
dflink=as.data.frame(linkincomefare)
names(dflink)=c('Fairtype','Purpose','Percent')
dflink_new = dflink %>% 
  mutate(Fairtype = recode(Fairtype, 'FALSE' = "Other")) #recoding FALSE to Other, meaning other type of fare
```

**Building the plot**

```{r barplotfarepurpose}
base  = ggplot(dflink_new, aes(x = reorder(Fairtype, Percent), y = Percent))
bars1 = base + geom_bar(stat = "identity")
bars2 = bars1 + facet_wrap( ~ Purpose ,nrow = 1)
bars3 = bars2 + coord_flip() + 
                scale_y_continuous(limits=c(0,80),
                                  labels=scales::unit_format(suffix = '%'))
bars4 = bars3 + geom_text(aes(y = Percent,
                                  label = paste0(round(Percent,1), '%')), 
                                  hjust= -0.15, 
                                  size = 2.5)
bars5 = bars4 + labs(title='What fare type is the most common given different purpose of trips?',
                     x ='Fare type',
                     caption = 'Source: Sound Transit 2015 Onboard Survey Data')
bars6 = bars5 + theme(axis.text.y = element_text(size=8, angle = 30),
                      axis.text.x = element_text(size=7),
                      plot.title = element_text(size = 12, hjust = 0.5),
                      plot.caption = element_text(hjust = 1, vjust = -1.2),
                      axis.title.y = element_text(size = 11, hjust = 0.5, vjust = 1.75),
                      axis.title.x = element_text(size = 11, hjust = 0.5, vjust = -1.25))
bars6
```


##################### MULTIVAR #################### 


In this section, we will explore the realationship among the probability of using cash for this trip, whether people have vehicles, the purpose of this trip, and frequency of using public transportation. 


**Preparing a dataset for regression** 

First, we need to make our dataset ready for regression. At the following code we select dependent variables and explanatory variables, then set the right format for each of the variables:

```{r multi transform binamiry variable_cash, message=FALSE, warning = FALSE}
#change names of explanatory variables
library(dplyr)
mydata1 <- rename(linkdata,'number_of_trips'='oneway_all_clean', 'vehicle'='hhveh_clean')

# transform explanatory varibles from factor to numeric, add "-1" in command because Factors in R are represented internally as integers and the first factor is 1.
mydata1$vehicle <- as.numeric(mydata1$vehicle)-1
mydata1$number_of_trips <- as.numeric(mydata1$number_of_trips)-1

# subset, select needed variables:
mydata2 <- mydata1 %>% mutate(cashlove = ifelse(faretype=="Cash", 1, 0), vehicle = ifelse(vehicle>0, 1, 0)) %>% select(cashlove, purpose, number_of_trips, vehicle) %>% filter(purpose != "#NULL!")
# set the correct type for explanatory variables
mydata2$cashlove <- factor(mydata2$cashlove)
mydata2$vehicle <- factor(mydata2$vehicle)
mydata2$purpose <- factor(mydata2$purpose) #reset the level of variable "purpose"
mydata2$number_of_trips <- as.numeric(mydata2$number_of_trips)
head(mydata2,5)
```

**Using the logit model** 

Here we use logit model since our dependent variable is a binary variable. Our reference group is people whose purpose of this trip is home-based non-work trip and don't have vehicles at home.
```{r logit regression, message=FALSE, warning=FALSE}
library(dotwhisker)
library(broom)
#just use purpose as explantary variable
mylogit1 <- glm(cashlove ~ purpose, data = mydata2, family = "binomial")
model1 = tidy(mylogit1) %>%   # we save the result as a tidy object and...
    mutate(model = "Model 1")
#use purpose and number of trips
mylogit2 <- glm(cashlove ~ purpose + number_of_trips, data = mydata2, family = "binomial")
model2 = tidy(mylogit2) %>%   # we save the result as a tidy object and...
    mutate(model = "Model 2")
#use all three explanatory varibales
mylogit3 <- glm(cashlove ~ purpose + number_of_trips + vehicle, data = mydata2, family = "binomial")
model3 = tidy(mylogit3) %>%   # we save the result as a tidy object and...
    mutate(model = "Model 3")

summary(mylogit3)
```

**Comparing three models**

```{r compare different models}
allModels=rbind(model1, model2, model3)

dwplot(allModels, vline = geom_vline(xintercept = 0, colour = "grey60", linetype = 2)) + 
  xlab("Coefficient Estimate") + 
  ylab("") +
  ggtitle("Predicting probability of using cash") +
  theme_bw()
```

From the plot above, we can see that compared with reference group, people whose purpose is home based work trip have less probability using cash, while people whose purpose is non-home based trip is more likely to use cash.
The more trips people have in each month, the less likelihood people would use cash. However, the difference on cash usage between people who have vehicles in home and people who don't is non-significant. 




